---
title: "01.1_create_training_material"
author: "Diego Aguilar-Ramirez"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
Follow the instructions here[https://support.rstudio.com/hc/en-us/articles/200526207-Using-RStudio-Projects]
```{r}
if (!require("pacman")) install.packages("pacman")
if(webshot::is_phantomjs_installed()==F)webshot::install_phantomjs()
pacman::p_load(gifski, ggplot2, rio, here, janitor, skimr, forcats, lubridate, tidyverse, gtsummary, rstatix, flextable, gt, webshot)
```

# 1. Create datasets used for lecture and practicals
## 1.1 Load STEPS data from WHO EpiInfo training material

For the course, I have manually exported from Access an `.xlsx` spreadsheet
called `STEPS_EpiInfo_Test.xlsx` that only contains the test dataset
as I had trouble importing the original `.mdb` file directly
into R.

```{r}
steps_test_raw <- import(here("data", "STEPS_EpiInfo_Test.xlsx"))
```
## Quick review of data with `skmir`
```{r}
skimr::skim(steps_test_raw)
```

## Clean and rename
Format and add descriptive names to columns
```{r}
steps_test <- steps_test_raw %>%
  
  janitor::clean_names() %>%
         # New name     # Old name
  rename(sex          = c1,
         age          = c3,
         curr_smkr    = t1,
         daily_smkr   = t2,
         curr_alc     = a1,
         freq_alc     = a2,
         bp_last_meas = h1,
         height_cm    = m3,
         weight_kg    = m4,
         sbp_1        = m11a,
         sbp_2        = m12a,
         sbp_3        = m13a) %>%
  
  mutate(sbp = (sbp_1+sbp_2+sbp_3)/3, # Here I simply calculate the mean of the 3 sbp measurements
         sex_fct = factor(sex, labels = c("Men", "Women"))) # Originally coded as 1 and 2
```



## Dataset for Lecture 1 needs to do the following:
###Add implausible values to SBP, only in women

```{r}
steps_test_lec1 <- steps_test %>%
  select(c(sex_fct,sbp,sex)) %>%
  drop_na() %>%
  add_row(sbp = 540, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 538, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 542, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 540, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 538, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 542, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 540, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 538, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 542, sex_fct = as.factor("Women")) %>%
  add_row(sbp = 540, sex_fct = as.factor("Women"))
```
### Make table with mean SBP by sex at different cleaning steps
```{r}
#Make table, then export as png, adding p values for differences in mean
step.table <- c()
model.names <- c("Before cleaning", "After dropping implausible values", "After dropping outliers")
sd.3 <- mean(na.omit(steps_test_lec1$sbp[steps_test_lec1$sbp<250])) + 3* sd(na.omit(steps_test_lec1$sbp[steps_test_lec1$sbp<250]))
df.list <- list(
  steps_test_lec1,
  filter(steps_test_lec1, sbp <250),
  filter(steps_test_lec1, sbp < sd.3)
)

for (i in 1:length(model.names)){
mod.name <- model.names[i]
mean_men <- stringr::str_glue("{round(mean(df.list[[i]]$sbp[df.list[[i]]$sex_fct=='Men'],na.rm=T),1)} ({round(sd(df.list[[i]]$sbp[df.list[[i]]$sex_fct=='Men'],na.rm=T), 1)})")
mean_women <- stringr::str_glue("{round(mean(df.list[[i]]$sbp[df.list[[i]]$sex_fct=='Women'],na.rm=T),1)} ({round(sd(df.list[[i]]$sbp[df.list[[i]]$sex_fct=='Women'],na.rm=T), 1)})")
diff <- stringr::str_glue("{round(mean(df.list[[i]]$sbp[df.list[[i]]$sex_fct=='Men'],na.rm=T) - mean(df.list[[i]]$sbp[df.list[[i]]$sex_fct=='Women'],na.rm=T), 1)} ({round(t.test(sbp ~ sex_fct,data=df.list[[i]])[[4]],1)[1]}, {round(t.test(sbp ~ sex_fct,data=df.list[[i]])[[4]],1)[2]})")
t_t <- round(t.test(sbp ~ sex_fct, data=df.list[[i]])[[3]], 3)
if(t_t<0.001)t_t <- "<0.001"
build.df <- data.frame(
  "Model"=mod.name,
  "Men"=mean_men,
  "Women"=mean_women,
  "Diff_in_means"=diff,
  "p-val"=t_t)

step.table <- rbind(step.table,build.df)
}
```

Export ste-pwise populated tables as png
```{r}
step.table.1 <-  
  step.table[1,] %>%
  flextable() %>%
  autofit() %>%
  add_header_row(
    top=TRUE,
    values=c("",
             "Mean (SD)",
             "",
             "",
             "")) %>%
  set_header_labels(
    Model = "",
    Men = "Men",
    Women = "Women",
    Diff_in_means = "Difference in means (95% CI)",
    p.val = "p-value") %>%
  merge_at(i=1, j=2:3,part="header") %>%
  flextable::align(align="center",j=c(2:3),part="header") %>%
  bold(i=1,bold=TRUE,part="header") %>%
  save_as_image(path = "figs/step-tbl-1.png")

step.table.2 <- 
  step.table[1:2,] %>%
  flextable() %>%
  autofit() %>%
  add_header_row(
    top=TRUE,
    values=c("",
             "Mean (SD)",
             "",
             "",
             "")) %>%
  set_header_labels(
    Model = "",
    Men = "Men",
    Women = "Women",
    Diff_in_means = "Difference in means (95% CI)",
    p.val = "p-value") %>%
  merge_at(i=1, j=2:3,part="header") %>%
  flextable::align(align="center",j=c(2:3),part="header") %>%
  bold(i=1,bold=TRUE,part="header") %>%
  save_as_image(path = "figs/step-tbl-2.png")

step.table.3 <- 
  step.table %>%
  flextable() %>%
  autofit() %>%
  add_header_row(
    top=TRUE,
    values=c("",
             "Mean (SD)",
             "",
             "",
             "")) %>%
  set_header_labels(
    Model = "",
    Men = "Men",
    Women = "Women",
    Diff_in_means = "Difference in means (95% CI)",
    p.val = "p-value") %>%
  merge_at(i=1, j=2:3,part="header") %>%
  flextable::align(align="center",j=c(2:3),part="header") %>%
  bold(i=1,bold=TRUE,part="header") %>%
  save_as_image(path = "figs/step-tbl-3.png")
```

### SBP in women with implausible values 
```{r}
plot1a.df <- steps_test_lec1 %>%
  
  filter(sex_fct=="Women")
  
  plot1a <- 
    ggplot(data = plot1a.df, mapping = aes(x = sbp)) +
      geom_histogram(binwidth = 5, color = "white", fill = "cadetblue") +
      #geom_density() +
      labs(
        title = "Systolic blood pressure in women",
        subtitle = stringr::str_glue("N={dim(plot1a.df)[1]}\nWHO STEPS training dataset"),
        x = "SBP",
        caption = stringr::str_glue("Mean SBP = {round(mean(plot1a.df$sbp),2)} (SD {round(sd(plot1a.df$sbp),2)}) mmHg")
      ) +
      theme_classic() #+
      #facet_wrap(~sex_fct)
  plot1a
```
Save
```{r}
ggsave(plot1a, filename = here("figs","plot1a.png"), width = 5, height = 4)
```

### SBP in women without implausible values
```{r}
plot1b.df <- plot1a.df %>%
  
  filter(sbp < 250)

  plot1b <- 
    ggplot(data = plot1b.df, mapping = aes(x = sbp)) +
      geom_histogram(binwidth = 5, color = "white", fill = "cadetblue") +
      #geom_density() +
      labs(
        title = "Systolic blood pressure in women",
        subtitle = stringr::str_glue("N={dim(plot1b.df)[1]}\nWHO STEPS training dataset"),
        x = "SBP",
        caption = stringr::str_glue("Mean SBP = {round(mean(plot1b.df$sbp),2)} (SD {round(sd(plot1b.df$sbp),2)}) mmHg")
      ) +
      theme_classic() #+
      #facet_wrap(~sex_fct)
  plot1b
```
Save
```{r}
ggsave(plot1b, filename = here("figs","plot1b.png"), width = 5, height = 4)
```

### SBP in men & women (histogram)
```{r}
plot2.df <- steps_test_lec1 %>%
  filter(sbp < 250)

  plot2<- 
    ggplot(data = plot2.df, mapping = aes(x = sbp)) +
      geom_histogram(binwidth = 5, color = "white", fill = "firebrick") +
      #geom_density() +
      labs(
        title = "Systolic blood pressure in men and women",
        subtitle = stringr::str_glue("N={dim(plot2.df)[1]}\nWHO STEPS training dataset"),
        x = "SBP",
        caption = stringr::str_glue("Mean SBP = {round(mean(plot2.df$sbp),2)} (SD {round(sd(plot2.df$sbp),2)}) mmHg")
      ) +
      theme_classic() +
      #facet_wrap(~sex_fct)
      geom_vline(
        xintercept = mean(plot2.df$sbp)+3*sd(plot2.df$sbp),
        color = "black",
        linetype = 2,
        alpha = 0.8) +
      geom_text(
        x = mean(plot2.df$sbp)+3.4*sd(plot2.df$sbp),
        y = 100,
        label = "3-SD"
      )
  
  plot2
```
Save
```{r}
ggsave(plot2, filename = here("figs","plot2.png"), width = 5, height = 4)
```

```{r}
  plot3<- 
    ggplot(data = plot2.df, mapping = aes(y = sbp)) +
      geom_boxplot(fill = "firebrick", width = 0.8) +
      #geom_density() +
      labs(
        title = "Systolic blood pressure in men and women",
        subtitle = stringr::str_glue("N={dim(plot2.df)[1]}\nWHO STEPS training dataset"),
        y = "SBP",
        x = "",
        caption = stringr::str_glue(
          "Mean SBP = {round(mean(plot2.df$sbp),2)} (SD {round(sd(plot2.df$sbp),2)}) mmHg\nMedian SBP = {round(quantile(plot2.df$sbp[3]),2)} (IQR {round(quantile(plot2.df$sbp[2]),2)}, {round(quantile(plot2.df$sbp[4]),2)}) mmHg")
      ) +
      theme_classic() +
      #facet_wrap(~sex_fct)
      geom_hline(
        yintercept = mean(plot2.df$sbp)+3*sd(plot2.df$sbp),
        color = "black",
        linetype = 2,
        alpha = 0.8) +
      geom_text(
        y = mean(plot2.df$sbp)+3.4*sd(plot2.df$sbp),
        x = -0.35,
        label = "3-SD") +
      scale_x_discrete(breaks = NULL)
  plot3
```
Save
```{r}
ggsave(plot3, filename = here("figs","plot3.png"), width = 5, height = 4)
```


## 1.2 Split dataset into 2
As part of the learning objectives is to merge datasets,
I will split the original data and preserve the ID column

```{r}
steps_test_1 <- dplyr::select(steps_test,one_of("ID", "H1", "M3", "M4", "M11A", "M12A", "M13A"))
steps_test_2 <- dplyr::select(steps_test,-c(H1:M13A))
```

## 1.3 Export two datasets
